---
title: "Obermayer_Assignment2"
author: "Alyssa Obermayer"
date: "9/14/2020"
output: html_document
---

## **R Markdown**Assignment 2**

## *Problem 1*

For problem 1 we will recreate the "Sit-and-Wait Forager" figure 2.7, from our 
textbook, in R using the same parameters as in the book. This model looks at 
sit-and-wait foraging of a semi-circle in front of a predator in two different 
habitats, wet and dry, and compares the average energy gain per prey item versus 
cutoff distance in each habitat. 

First, we input the parameters for the model
```{r}
# Initial conditions and parameters from the book

e = 10                             #Energy per prey item
ew = 0.1                           #Energy spent waiting
ep = 1                             #Energy spent pursuing
v = 0.5                            #Running speed of predator
rc = seq(0.005, 4.000, by = 0.005) #Cutoff distances
aw = 0.02                          #Prey arrival rate wet habitat
ad = 0.005                         #Prey arrival rate dry habitat
```

Next, we will generate equations to analyze the average energy gain per prey item 
in both the wet and dry habitats.
```{r}
#Wet habitat analysis
#Equation for time spent waiting in the wet habitat
tww <- 2/(aw * pi * (rc^2))

#Equation for average pursuit time
tp <- (4/3) * (rc/v)

#Equation for total average time in wet habitat
tiw <- (tww + tp)

#Equation for average energy per item in wet habitat
eiw <- (e) - (2 * ((ew)/(aw * pi * (rc^2)))) - ((4/3) * ((ep * rc)/(v)))

#Equation for average energy gain per item in wet habitat
etw <- (eiw)/(tiw)

#Dry habitat analysis
#Equation for time spent waiting in the dry habitat
twd <- 2/(ad * pi * (rc^2))

#Equation for total average time in wet habitat
tid <- (twd + tp)

#Equation for average energy per item in wet habitat
eid <- (e) - (2 * ((ew)/(ad * pi * (rc^2)))) - ((4/3) * ((ep * rc)/(v)))

#Equation for average energy gain per item in wet habitat
etd <- (eid)/(tid)
```

Finally, with the quantitative information determined above we can generate the plot.
```{r}
#Sit and wait foraging plot
plot(etw~rc,
     type = "l",
     xlab = "Cutoff Distance (m)",
     ylab = "Energy/Time (j/s)",
     main = "Sit-and-Wait Forager",
     col = "cadetblue")
lines(etd~rc,
      col = "burlywood4")
```

This plot represents two curves. The blue curve represents the optimal foraging 
distance when in the wet habitat and the brown curve represents the dry habitat. 
The curve of the wet habitat peaks higher and at a shorter cutoff distance than 
the curve of the dry habitat. When generating the plot, we took into account 
that the wet habitat has a higher prey arrival rate than the dry habitat, so 
the predator would not have to wait as long for their prey when foraging, nor 
travel as far to have an optimal outcome. In contrast, the dry habitat has a 
lower prey arrival rate, so the predator has to wait longer for the prey, and 
travel farther to obtain prey for an optimal outcome. This shows that foraging 
in the wet habitat is optimal when compared to the dry habitat, because it 
results in a higher average energy intake per prey item over time when foraging.

This figure could also inform us of the possible threshold at which the predator 
might decide to accept or ignore the prey they come into contact with. If the 
prey item is too far from the predator, it may not be worth the effort to try 
and obtain it. The threshold determined by this graph would be the best 
compromise to minimize wait time and minimize pursuit time to maximize the 
energy gains, which would be considered the optimal cutoff distance and thus 
part of the optimal foraging strategy for the predator. This can be biologically 
informative as it would tell us which habitat might be better suited for 
foraging for the species in question.